Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-21 Manuscript/manuscript.Rmd
somehow a lot of my edits (mostly methods section) were overwritten, I have reincluded them

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangements of higher-order chromatin architecture [@phillips2013chromatin]. Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long-range chromatin contacts [@zuin2014cohesin]. Other regulatory elements like particular histone modifications such as H3K36 and H3K27 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions [??? Better to mention those that came up in your analysis, otherwise, it is unnecessary information] [@rao20143d]. These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the TAD boundaries, allowing us to better understand what leads to their formation.

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not [@mourad2016computational; @hong2017computational; @huang2015predicting] [??? Do not understand this sentence]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data, such as heavily imbalanced number ot boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions mostly constitute as resampling techniques, in which random sampling is used to either replicate or removed data points in order to create a more balanced dataset [??? REFS]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study we focus our attention toward data level solutions so that all results may be classifier independent, and thus, more easily comparable.

@@

Replicate Hi-C data for the GM12878 cell line were downloaded from GEO GSE63525 (Experiments HIC001 to HIC018; produced using the MbolI restriction enzyme) [@rao20143d]. The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained. Genomic annotation data (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table S1).

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. Returning to the 10 kb resolution example, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

To accommodate for the uncertainty in the location of TAD boundary points, they were flanked on both sides by regions of the size of the resolution of the data (Figure 1). For example, returning to the 10 kb resolution example, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb.

The response vector $Y$ was formed by genomic regions (bins) defined by the resolution of the data. Genomic regions flanking TAD boundaries were labeled as $Y=1$; otherwise, the regions were labeled as not containing a boundary ($Y=0$).

Using the sets of genomic features, we created three types of predictors (Figure 2):

1. **Overlap Counts (OC)**: For each genomic bin, the total number of overlaps with genomic features was calculated. In the case of no overlaps the value is set as 0.

2. **Overlap Percent (OP)**: For each genomic bin, the percent overlap was defined as the total width of features overlapping a bin divided by the bin width. Values for bins with no overlaps were set to be 0.

3. **Distance**: For each genomic bin, the log [??? didn't you test raw distance first?] [??? log10?] of the distance (in bases) from the center of the bin to the center of the nearest genomic feature region was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figure 1).

@@

@@

In order to evaluate the problem of imbalanced classes, we then evaluated three re-sampling techniques.

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE. We chose the parameters that would create perfectly balanced classes. The `DMwR` v.3.4.2 R package was used to perform SMOTE.

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` v.3.4.2 [??? Is it correct? Same version as DMwR?] R packagewas used to perform ROSE.

@@

A classification algorithm in the form of a random forest was performed and validated using the same test [??? Training?] set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. The evaluation metrics listed above are all considered to be single threshold measures. That is, they are defined for individual thresholds that separated class prediction. Thererfore, they are unable to give an overview of the potential range of performance as thresholds vary. In order to evaluate the trade-off between precision and recall, we also considered a threshold-free measure in form of the area under the precision recall curve (AUPRC).

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest model. For each tree, the prediction accuracy on the out-of-bag portion of the data is recorded. Then the same was done after permuting each predictor variable. The difference between the two accuracies were then averaged over all trees, and normalized by the standard error [@kuhn2012caret]. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models within each set of predictor types. Here, random forest models were used to assign weights to a set features of a given size. The optimal number of features was determined by the AUPRC of the corresponding model. [??? Rewrite this paragraph as I put some disjoint sentences]

>>>>

Identifying the mechanisms underlying the formation of TADs are a complex and active area of research. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the TAD boundaries. [??? Better to mention those that came up in your analysis, otherwise, it is unnecessary information; !!!editted paragraph] Current methods rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries [@mourad2016computational; @hong2017computational; @huang2015predicting] [??? Do not understand this sentence; !!!editted sentence]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data, such as heavily imbalanced number ot boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

@@

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions mostly constitute as resampling techniques, in which random sampling is used to either replicate or removed data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data] [??? REFS; !!!added references]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study we focus our attention toward data level solutions so that all results may be classifier independent, and thus, more easily comparable.

@@

Replicate Hi-C data for the GM12878, K562, ... cell lines were downloaded from GEO GSE63525 (Experiments HIC001 to HIC018, HIC069 to HIC 074, ... respectively; produced using the MbolI restriction enzyme) [@rao20143d]. The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the Arrowhead algorithm [@durand2016juicebox]. The start and end coordinates of all TADs were concatenated, sorted, and unique TAD boundary points were obtained. Genomic annotation data consisting of various chromatin segmentation states, histone modifications, and transcription factor binding sites (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table S1).

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. Returning to the 10 kb resolution example, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

## GENOMIC ANNOTATION DATA

Genomic annotation data (GRCh37/hg19 human genome assembly) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 1). The annotations consisted of a variety of different chromatin states, histone marks, and transcription factor binding sites using ChIP-seq peak experiments. Each genomic annotation was represented by a set of regions (their genomic coordinates) annotated as having a functional property or biological significance.

## ENSEMBLE FRAMEWORK

An ensemble of models was established for the purposes of systematically comparing the performances of different parameters in the framework. Each system in the framework is assumed to be composed of a model constructed from a particular data resolution, with a particular predictor type for the feature space (Figure 2).

@@

### FEATURE ENGINEERING AND REDUCTION USING GENOMIC ANNOTATIONS

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries (Figure 3):

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances where a bin overlapped with genomic annotation regions (features) was calculated. In the case of no overlaps the value is set as 0.

2. **Overlap Percent (OP)**: For each genomic bin, the percent overlap between the feature width and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with the genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 10 kb, 25 kb, 50 kb, or 100 kb). Values for bins with no overlaps were set to be 0.

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance (in bases) from the center of the bin to the center of the nearest region of a respective genomic annotation region was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figure 1)

To assess the predictive performance for each predictor type, we performed individual random forest classification algorithms utilizing each set predictor types as the feature space, in addition to a model with all combined types. Precision-Recall (PR) Curves, and subsequent area under the PR curves (AUPRC), were used to evaluate which predictor type contributed to the best model performance. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models within each set of predictor types. In doing so, random forest models were used to assign weights to a set features of a given size. The optimal number of features was determined by the AUPRC of the corresponding model in order to reduce the feature space accordingly.

@@

In order to evaluate the problem of imbalanced classes, we then evaluated 4 re-sampling techniques (in addition to no re-sampling).

@@

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE. We chose the parameters that would create perfectly balanced classes. The `DMwR` package in R was used to perform SMOTE (version 0.4.1).

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` package [??? Is it correct? Same version as DMwR?; !!!fixed] in R used to perform ROSE (version 0.0.3).

@@

A classification algorithm in the form of a random forest was performed on the various resampled and reduced training sets and validated using the same test [??? Training?; !!!clarified; different training sets due to the different resampling schemes, but validated using the same holdout set] set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest model. For each tree, the prediction accuracy on the out-of-bag portion of the data was recorded. Then the same was done after permuting each predictor variable. The difference between the two accuracies was then averaged over all trees, and normalized by the standard error [@kuhn2012caret].
